---
title: "Project 1 Checkpoint 4"
author: "Mira Griffin, Eloise Melendez, Chloe Schurter"
format: html
embed-resources: true
code-fold: true
theme: lux
---



## Do certain months of the year tend to have a higher amount of suspected measles cases, for different countries? 

```{r} 
#| include: false
library(tidyverse)
library(ggridges)
library(gt)

cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
```

```{r}
#| warning: false
top_pop <- cases_year |>
  group_by(region) |>
  slice_max(order_by = total_population, n = 1, with_ties = FALSE)
  
top_pop_month <- left_join(top_pop, cases_month, by = "country")

top_pop_month$month <- factor(top_pop_month$month, 
                              levels = 1:12,
                              labels = c("Jan", "Feb", "Mar","Apr",                                          "May","Jun", "Jul", "Aug",                                          "Sep", "Oct", "Nov",                                               "Dec"))

ggplot(data = top_pop_month,
       mapping = aes(x = measles_suspect,
                     y = month,
                     fill = country)) +
  geom_density_ridges(alpha = 0.5) +
  labs(x = "Number of Suspected Measles Cases", 
       y = "Month", 
       fill = "Country", 
       title = "Monthly Distributions of Suspected Measles Cases Across Countries",
       subtitle = "Data from 2012-2025") +
  scale_x_continuous(breaks = seq(0, 30000, 5000)) +
  scale_fill_brewer(palette = "Spectral") +
  theme_light() +
  theme(text = element_text(family = "Times", size = 12))
```

## Is the ratio of measles cases to rubella cases per year different depending on the country? (mutate)

```{r}
#| include: false 
measles_prop_year <- cases_year |> 
  group_by(country, year) |> 
  mutate(prop_tot_measles_per_year = ifelse(
      measles_total + rubella_total == 0,
      NA,
      measles_total / (measles_total + rubella_total)
    )) |>
  select(region, country, year, prop_tot_measles_per_year)

measles_prop_tot <- measles_prop_year |>
  group_by(country) |>
  summarize(avg_prop_tot_measles = 
              ifelse(all(is.na(prop_tot_measles_per_year)),
                     NA,
                     mean(prop_tot_measles_per_year, 
                          na.rm = TRUE)), .groups = "drop")
measles_prop <- measles_prop_year |> left_join(measles_prop_tot,
                                               by = "country")
measles_prop

```
```{r}
#| include: false 
top_pop = cases_year |> 
  group_by(region) |>
  slice_max(order_by = total_population, n = 14, with_ties = FALSE)

top_pop <- top_pop[-c(26, 27, 28, 69, 70), ]
```


```{r}
#| include: false 
measles_prop_year <- top_pop |>
  group_by(country, year) |> 
  mutate(prop_tot_measles_per_year = ifelse(
      measles_total + rubella_total == 0,
      NA,
      measles_total / (measles_total + rubella_total)
    )) |>
  select(region, country, year, prop_tot_measles_per_year)

measles_prop_year

measles_prop_tot <- measles_prop_year |>
  group_by(country) |>
  summarize(avg_prop_tot_measles = 
              ifelse(all(is.na(prop_tot_measles_per_year)),
                     NA,
                     mean(prop_tot_measles_per_year, 
                          na.rm = TRUE)), .groups = "drop")
measles_prop <- measles_prop_year |> left_join(measles_prop_tot,
                                               by = "country")
measles_prop

```

```{r}
ggplot(data = measles_prop,
      mapping = aes(x = year,
                    y = prop_tot_measles_per_year,
                    color = country)) +
 geom_point() +
 theme_light() +
 ggtitle("Ratio of Measles Cases out of Measles and Rubella Cases in Countries from Each Region with the Highest Population"
 ) +
 labs(x = "Year", y = "Proportion of Measles Cases")
```



## Multiple Linear Regression Model for Measles Proportion by Country and Year 
```{r}
#| include: false

cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')

measles_prop_year <- cases_year |> 
  group_by(country, year) |> 
  mutate(prop_tot_measles_per_year = 
    measles_total / (measles_total + rubella_total)
  ) |>
  select(country, year, prop_tot_measles_per_year) |>
  ungroup()

```


```{r} 
#| include: false
measles_model <- lm(prop_tot_measles_per_year ~ country + year, 
                    data = measles_prop_year)
summary(measles_model) 
```

```{r}
#| include: false 

model_df <- as.data.frame(
  summary(measles_model)$coefficients, check.names = FALSE) |>
  rownames_to_column("term") |>
  filter (term != "Intercept") |>
  arrange(desc(abs(Estimate))) |>
  mutate(
    term = gsub("country", "", term),
    Estimate = round(Estimate, 3),
    `Std. Error` = round (`Std. Error`, 3),
    `t value` = round (`t value`, 2),
    `p value` = round (`Pr(>|t|)`, 3)) |>
  select(c("term","Estimate", "Std. Error", "t value", "p value"))
model_df
```

```{r}

model_table <- model_df |>
  gt() |>
  tab_header(
    title = md("Multiple Linear Regression Model for Measles Proportion by Country and Year")
  ) |>
  tab_caption(
    caption = md("Estimates of differences in measles proportion (relative to the reference country), in descending order")) |>
  
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  fmt_percent( columns = Estimate) |>
  cols_label(
    term = "Country",
    Estimate = "Estimate",
    `Std. Error` = "Std Error",
    `t value` = "t-value",
    `p value` = "p-value"
  ) |>
  fmt_number(columns = -c(term, Estimate),
             decimals = 4
  )
  
model_table  

```



This model is statistically significant, and it can be concluded that at
least one country or year effect differs in the proportion of measles to rubella cases. 
(F = 7.047, df = 169, p value = 2.2e-16). About 42.08% of the variation in proportions
of measles to rubella cases can be explained by country and year (R^2 = 0.4208).
Year was a significant positive predictor in proportion of measles cases over many years. 
(estimate = 9.181e-03, p value = 4.14e-13)

We can see that the top 5 countries with the largest absolute difference
in measles proportions relative to the reference country are
Poland (largest), Botswana, Zimbabwe, Eswatini and Madagascar. For example, Poland
has the largest difference in percent (69.90% less than the reference country)

