---
title: "Project 1 Checkpoint 2"
author: "Mira Griffin, Eloise Melendez, Chloe Schurter"
format: html
---
```{r}
#| echo: false
#| include: false

library(tidyverse)

cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')

```

## Do certain months of the year tend to have a higher amount of suspected measles cases, for different countries? (custom function/iteration)


```{r}

summ_sus_measles_per_month <- function(country_name){
  country_data <- cases_month |> 
    filter(country == country_name)
  
  summ_sus_measles <- tibble()
  
  for(mon in 1:12) {
    mon_data <- country_data |> 
      filter(month == mon)
    
    summ_sus_measles <- bind_rows(summ_sus_measles, tibble(
      country = country_name,
      month = month.name[mon],
      mean = mean(mon_data$measles_suspect, na.rm = TRUE),
      SD = sd(mon_data$measles_suspect, na.rm = TRUE)
    ))
  }
  
  return(summ_sus_measles)
}

summ_sus_measles_per_month("Algeria")
summ_sus_measles_per_month("Bulgaria")
```

## Is the ratio of measles cases to rubella cases per year different depending on the country? (mutate)
```{r}
measles_prop_year <- cases_year |> 
  group_by(country, year) |> 
  mutate(prop_tot_measles_per_year =                                             ifelse(
      measles_total + rubella_total == 0,
      NA,
      measles_total / (measles_total + rubella_total)
    )) |>
  select(country, year, prop_tot_measles_per_year)

measles_prop_tot <- measles_prop_year |>
  group_by(country) |>
  summarize(avg_prop_tot_measles = 
              ifelse(all(is.na(prop_tot_measles_per_year)),
                     NA,
                     mean(prop_tot_measles_per_year, 
                          na.rm = TRUE)), .groups = "drop")
measles_prop <- measles_prop_year |> left_join(measles_prop_tot,
                                               by = "country")
measles_prop
```


## Does the proportion of discarded measles cases to suspected measles cases decrease from year to year? (iteration over years, proportion column, dummy yes/no variable)

```{r}
cases_year_prop_discarded <- cases_year |>
  arrange(year) |>
  mutate (prop_discarded_to_total = discarded_cases / total_suspected_measles_rubella_cases) 


prop <- cases_year_prop_discarded$prop_discarded_to_total
cases_year_prop_discarded$decrease <- map_int (seq_along(prop), function(x){
  if (x == 1){
    return (NA)
  }
    
  ifelse(prop[x] < prop[x-1], 1, 0) 
})


```

## Do certain regions have higher measles incident rates per 1,000,000 people, on average? (columns are regions, average incident rate on rows, iterate over regions)

cleaned <- cases_year |> 
  select(region, measles_incidence_rate_per_1000000_total_population)

region_avg <- data.frame(region = unique(cleaned$region),
                         avg_incidence = NA)

for (i in 1:nrow(region_avg)) {
  r <- region_avg$region[i]
  region_avg$avg_incidence[i] <- 
    mean(cleaned$measles_incidence_rate_per_1000000_total_population[cleaned$region == r], 
         na.rm = TRUE)
}

region_avg

